Assessing the Accessibility of Protein Tunnels using Motion Planning

### 2018-02-24 paper.tex
spelling 2

\begin{abstract}
Chemical interactions between proteins and other molecules (ligands) take place in specific locations, called active sites, that are often deeply buried inside the protein structure.
These active sites are accessible through one or more void paths, called tunnels.
Nowadays, tunnels are mostly computed using Voronoi diagrams which approximate the ligand by a bounding sphere and 
the traversability of the tunnels is determined based only on the radial bottlenecks.
This can lead to imprecise results as the shape of the ligand is not considered.
To address this problem, we define the computation of the transportation path of non-spherical ligands including their conformation changes (i.e., when the atoms of the ligand move relatively to each other) as a motion planning task in the high-dimensional configuration space.
We propose a modification of the Rapidly Exploring Random Tree planner for trajectory computations of flexible ligands.
To cope with the narrow passage problem, the random samples are generated around a tunnel detected by a Voronoi-based approach.
%which serves us for better localization of the searched space.
The flexibility of the ligand is modeled using a fixed set of known conformations. 
%which does not increase the number of dimensions of the configuration space.
The proposed method has been evaluated in a virtual screening scenario to estimate the traversability of tunnels for several ligands.
\end{abstract}

@@

The analysis of the traversability of tunnels by flexible ligands can be formulated as a motion planning problem and solved by sampling-based planners.
The sampling-based planners randomly sample the configuration space of the robot (ligand) and connect the collision-free samples into a graph structure (roadmap).
Rapidly Exploring Random Tree (RRT)~\cite{lavalleRRT} is one of the most used sampling-based planners.
Besides many applications in robotics, the sampling-based planners have been already used in biochemistry as well, 
e.g., to study loop motions~\cite{cortes2004geometric}, to identify binding sites~\cite{bayazit2001ligand}, for exit
pathway detection~\cite{cortes2010simulating,cortes2005path}, for tunnel detection~\cite{vonasek2017tunnel,vonasek2016application}, and especially for protein folding~\cite{raveh2009rapid,amato2002using}. %novinskaya2015improving

@@

In the case of RRT-based planners, increasing the probability of sampling in difficult regions does not need to bring any advantage
if the growth of the tree is blocked by the obstacles.
%The RRT-based planners cannot cope with the narrow passage problem by increasing the probability of sampling in the difficult regions, as the growth of the configuration tree towards the random samples may be blocked by obstacles. %~\cite{vonasekphd}.
To prevent this blocking, DD-RRT~\cite{yershovaDDRRT} limits the selection of nodes for the expansion to a small ball. 
The radius of this ball is set to infinity for new nodes, and decreases to a predefined radius if the node cannot be successfully expanded.
An automatic adaptation of this radius was proposed in the ADD-RRT~\cite{jailletADRRT}.
%The disadvantage of DD-RRT is the sensitivity to the predefined radius.
%Therefore, the authors of ADD-RRT~\cite{jailletADRRT} proposed to adapt the ball radius according to the success rate of the expansion.
To attract the tree towards a given region, random samples have to be generated there only if the tree can expand to this region.
In~\cite{kardossRRTKK}, the probability of sampling is increased in several waypoints close to narrow passages, but without specifying how to find these waypoints.
The generalization of~\cite{kardossRRTKK} is the guided sampling, where the whole path is used to generate samples in the configuration space.
The guiding path can be computed in the workspace~\cite{vonasek2009rrt}, or iteratively refined in the configuration space based on the solution of a relaxed version of the problem~\cite{bayazitIRC}.

@@

It has been observed, that the tunnels detected in protein conformations without ligands (provided by MD simulations without ligands or by
protein conformation sampling) are rather narrow. 
However, in protein dynamics with ligand, the tunnels can move, merge and adapt to the shape of the passing ligand and vice versa.
%Therefore, it can be expected that the detected tunnels are rather narrow.
%Due to the protein dynamics, the tunnels move and merge, and their bottleneck can change in each time step.
%The tunnels can adapt their shape to the passing ligands and vice versa.
In consequence, even very narrow tunnels, being too small for the passage of a given ligand-approximating sphere, can potentially serve as the transportation path for the ligand.
Therefore, also the existing Voronoi-based methods enable the user to decrease the size of the approximating sphere which simulates the shrinking of the ligand size.
In our solution, we simulate this by shrinking the atom radii of the ligand, which is used also by~\cite{cortes2010simulating,guieysse2008structure}.

MD simulations of ligands passing through proteins further show that the ligands can be highly flexible.
The possible conformations of a ligand are determined by the torsion and dihedral angles between its atoms. 
The low-energy conformations are preferred.
%the ligand can contain several torsion and dihedral angles between its atoms which determine possible ligand conformations.
Therefore, the ligand flexibility can be modeled using only a predefined set of low-energy conformations.
By using the set of predefined conformations, it is not necessary to model the additional degrees of freedom, so 
the dimension of the configuration space is not increased as in~\cite{cortes2010simulating}.
%So the number of dimensions of the configuration space does not depend on the DOF of the ligand as in~\cite{cortes2010simulating}.
%The libraries containing possible ligand conformations are used also in different types of calculations, e.g., in~\cite{kellogg}.

@@

The proposed algorithm was verified in two scenarios: on a sequence of frames of MD simulation and on a set of protein conformations provided 
the conformation sampling approach.
%by the tCONCOORD tool~\cite{seeliger2007geometry}.

@@

The methods \RA\ and \RB\ differ in the expansion step (lines \ref{alg::expa}--\ref{alg::expb} in Alg.~\ref{alg::expand}).
In these lines, the \RB\ method employs the retraction step in order to find a new collision-free configuration $\qnew$ that approaches $\qrand$.
The retraction step is realized by sampling the vicinity of the configuration $\qnear$ in the distance $0.05$~\AA\ using 20 samples.
A collision-free sample that maximally approaches $\qrand$ is selected and added to the tree, and the retraction continues from this sample.
The retraction step is repeated 5 times.
The \RB\ method can therefore expand the tree by at most 5 new nodes per ligand conformation, while \RA\ extends the tree only by one new node.
The larger number of nodes in \RB\ method is not problematic, as the KD-trees were used for the nearest-neighbor search and the time to solve these queries is significantly smaller than the computing collision detection.
The parameters of the methods were selected in such a way that the number of collision detection queries in each expansion step is the same.
For each conformation and tested scale, the \RA\ method checks $m=100$ candidate configurations for collision, and \RB\ checks $5 \times 20$ configurations for collision.
%collision-detection queries.
Other parameters of the \RB\ planner were the same as for \RA.
The planning was terminated after $\Imax$ iterations or if the tree approached the active site to the distance less than 2~\AA\ (i.e., the 
geometric center of the ligand was closer than 2~\AA\ from the active site).

@@

The traversability rate is shown in Tab.~\ref{tab::main}. 
Only the traversability of the main tunnel is shown, as no trajectory was found in the side tunnel for most of the tested ligands.
The only exception is the ligand m037t, which was the smallest tested one (consisting of 8 atoms), and therefore it could fit even inside the side tunnel.
The traversability rate is higher for the most scaled-down ligands ($\smin=0.5$) and it decreases with the increasing scale.
For the DCP ligand, the trajectories were detected in 18~\% of frames for scale $\smin=0.8$ and the traversability of this ligand
to the active site was approved by a separate MD simulation~\cite{marques2017catalytic}.
%For the scale $\smin=0.8$, the most traversable frames were detected for ligands DCP (18~\%), m037t (45~\%) and m038t (30~\%).
%The traversability of the DCP ligand to active site was approved by a separate MD simulation~\cite{marques2017catalytic}.
Based on the success rate, also other ligands (e.g., m037t with traversability rate 45~\% at $\smin$=0.8) seem to be promising.
However, their binding was not yet approved by dedicated MD simulations.
Most of the used 50 ligand conformations where used in the planning, which is shown in Tab.~\ref{tab::m040c}.

@@

\begin{table}[t]
\caption{\label{tab::m040c}
\small
Examples of elongated/packed conformation for m040 (1,5-Dichloropentane).
The table shows the average number of conformations (out of 50) used in trajectories reaching the active site in the 4E46 protein.
}
\centering
{\footnotesize
%\def\arraystretch{0.9}
\begin{tabular}{cc}
\rotatebox{0}{\hskip 5pt Long} \\ \includegraphics[width=0.07\textwidth]{fig/m040-conf1} \\
\rotatebox{0}{Packed} \\ \includegraphics[width=0.0375\textwidth]{fig/m040-conf2}
\end{tabular}
}
{\scriptsize
\def\arraystretch{0.9}
\input usedc.tex
}
\end{table}

@@

%The case was the analysis of Leukotriene A4 hydrolase (PDB ID 4L2L).
The task of the second analysis was to find the trajectory for the ligand 4-(4-benzylphenyl)-1,3-thiazol-2-amine into the active
site of A4 hydrolase (PDB ID 4L2L, 9666 atoms).
%This ligand is a known inhibitor of the protein 4L2L.
%Based on the location of inhibitor inside the protein we assume that the inhibitor is able 
The tCONCOORD tool~\cite{seeliger2007geometry} was used to generate 1000 conformations of the protein 4L2L.
The initial PDB structure was minimized by GROMACS 4.0.7
%\cite{hess2008gromacs} 
 using the input parameters recommended on the example page tCONCOORD~\cite{tconcoord}.
The protein conformations were generated using the default settings for tCONCOORD in the PERT mode. % which only perturbs the starting configuration of atoms instead of complete randomization.
The tunnels leading to the active site defined by residues 134 (Alanine) and 311 (Phenylalanine) were computed by CAVER 3.0, using the probe 0.9~\AA.
All 1000 conformations contained at least 5 tunnels, more than 900 conformations contained 9 tunnels.
The ordering of the detected tunnels slightly differs in different conformations.
Therefore, the ligand traversability towards the active site was computed for the first three widest tunnels (with average bottlenecks 1.22~\AA, 1.20~\AA\, and 1.19~\AA\ with the average lengths 22.3~\AA, 25.3~\AA\, and 28.4~\AA, respectively).
The conformations of the ligand were prepared in the same way as in the previous section.
From the available $|\L|=736$ low-energy conformations, 100 were selected using the `S' strategy.

@@

As the tested ligand is a known inhibitor of the protein 4L2L~\cite{marques2017enzyme} we assume that the ligand can reach the active site.
Due to large size of the ligand (33 atoms), low scale-down factors $\smin\in\{0.4,0.5,0.6\}$ were used and
the active site was considered as reached if the distance to the geometric center of the ligand was less than 5~\AA.
All parameters of the \RA\ were set as in the previous experiment, except $\Imax=100\cdot10^3$.
The proposed planner was able to find trajectories even for this large ligand inside the long tunnels, which is
shown by the traversability rates (Tab.~\ref{tab::4l2l}).
Example of the solution is depicted in Fig.~\ref{fig::inhibitor}.

>>>>

\begin{abstract}
Chemical interactions between proteins and other molecules (ligands) take place in specific locations, called active sites, that are often deeply buried inside the protein structure.
These active sites are accessible through one or more void paths, called tunnels.
Nowadays, tunnels are mostly computed using Voronoi diagrams which approximate the ligand by a bounding sphere and 
the traversability of the tunnels is determined based only on the radial bottlenecks.
This can lead to imprecise results as the shape of the ligand is not considered.
To address this problem, we define the computation of the transportation path of non-spherical ligands including their conformation changes (i.e., when the atoms of the ligand move relatively to each other) as a motion planning task in the high-dimensional configuration space.
We propose a modification of the Rapidly Exploring Random Tree planner for trajectory computation of flexible ligands.
To cope with the narrow passage problem, the random samples are generated around a tunnel detected by a Voronoi-based approach.
%which serves us for better localization of the searched space.
The flexibility of the ligand is modeled using a fixed set of known conformations. 
%which does not increase the number of dimensions of the configuration space.
The proposed method has been evaluated in a virtual screening scenario to estimate the traversability of tunnels for several ligands.
\end{abstract}

@@

The analysis of the traversability of tunnels by flexible ligands can be formulated as a motion planning problem and solved by sampling-based planners.
The sampling-based planners randomly sample the configuration space of the robot (ligand) and connect the collision-free samples into a graph structure (roadmap).
Rapidly Exploring Random Tree (RRT)~\cite{lavalleRRT} is one of the most used sampling-based planners.
Besides many applications in robotics, the sampling-based planners have been already used in biochemistry as well, 
e.g., to study loop motions~\cite{cortes2004geometric}, to identify binding sites~\cite{bayazit2001ligand}, for exit
pathway detection~\cite{cortes2010simulating,cortes2005path}, for tunnel detection~\cite{vonasek2017tunnel}, 
        and especially for protein folding~\cite{raveh2009rapid,amato2002using}. %novinskaya2015improving

@@

In the case of RRT-based planners, increasing the probability of sampling in difficult regions does not need to bring any advantage
if the growth of the tree is blocked by the obstacles.
%The RRT-based planners cannot cope with the narrow passage problem by increasing the probability of sampling in the difficult regions, as the growth of the configuration tree towards the random samples may be blocked by obstacles. %~\cite{vonasekphd}.
To prevent this blocking, DD-RRT~\cite{yershovaDDRRT} limits the selection of nodes for the expansion to a small ball. 
The radius of this ball is set to infinity for new nodes, and decreases to a predefined radius if the node cannot be successfully expanded.
An automatic adaptation of this radius was proposed in the ADD-RRT~\cite{jailletADRRT}.
%The disadvantage of DD-RRT is the sensitivity to the predefined radius.
%Therefore, the authors of ADD-RRT~\cite{jailletADRRT} proposed to adapt the ball radius according to the success rate of the expansion.
To attract the tree towards a given region, random samples have to be generated there only if the tree can expand to this region.
In~\cite{kardossRRTKK}, the probability of sampling is increased in several waypoints close to narrow passages, but without specifying how to find these waypoints.
The generalization of~\cite{kardossRRTKK} is the guided sampling, where a sequence of waypoints (a guiding path) is used to generate samples in the configuration space.
The guiding path can be computed in the workspace~\cite{vonasek2009rrt}, or iteratively refined in the configuration space based on the solution of a relaxed version of the problem~\cite{bayazitIRC}.

@@

It has been observed, that the tunnels detected in protein conformations without ligands (provided by MD simulations without ligands or by
protein conformation sampling) are rather narrow. 
However, MD simulations of proteins with ligands have shown that the tunnels can move, merge and adapt to the shape of the passing ligand and vice versa.
%Therefore, it can be expected that the detected tunnels are rather narrow.
%Due to the protein dynamics, the tunnels move and merge, and their bottleneck can change in each time step.
%The tunnels can adapt their shape to the passing ligands and vice versa.
In consequence, even very narrow tunnels, being too small for the passage of a given ligand-approximating sphere, can potentially serve as the transportation path for the ligand.
Therefore, also the existing Voronoi-based methods enable the user to decrease the size of the approximating sphere which simulates the shrinking of the ligand size.
In our solution, we simulate this by shrinking the atom radii of the ligand, which is used also by~\cite{cortes2010simulating,guieysse2008structure}.

MD simulations of ligands passing through proteins show that the ligands can be highly flexible.
All possible conformations of a ligand are determined by the torsion and dihedral angles between its atoms, but the low-energy conformations are preferred.
%the ligand can contain several torsion and dihedral angles between its atoms which determine possible ligand conformations.
Therefore, the ligand flexibility can be modeled using only a predefined set of low-energy conformations.
By using the set of predefined conformations, it is not necessary to model the additional degrees of freedom, so 
the dimension of the configuration space is not increased as in~\cite{cortes2010simulating}.
%So the number of dimensions of the configuration space does not depend on the DOF of the ligand as in~\cite{cortes2010simulating}.
%The libraries containing possible ligand conformations are used also in different types of calculations, e.g., in~\cite{kellogg}.

@@

The proposed algorithm was verified in two scenarios: on a sequence of frames of MD simulation and on a set of protein conformations 
provided by the conformation sampling approach.
%by the tCONCOORD tool~\cite{seeliger2007geometry}.

@@

The methods \RA\ and \RB\ differ in the expansion step (lines \ref{alg::expa}--\ref{alg::expb} in Alg.~\ref{alg::expand}).
In these lines, the \RB\ method employs the retraction step in order to find a new collision-free configuration $\qnew$ that approaches $\qrand$.
The retraction step is realized by sampling the vicinity of the configuration $\qnear$ in the distance $0.05$~\AA\ using 20 samples.
A collision-free sample that maximally approaches $\qrand$ is selected and added to the tree, and the retraction continues from this sample.
The retraction step is repeated 5 times.
The \RB\ method can therefore expand the tree by at most 5 new nodes per ligand conformation, while \RA\ extends the tree only by one new node.
The larger number of nodes in \RB\ method is not problematic, as the KD-trees were used for the nearest-neighbor search and the time to solve these queries is significantly smaller than the collision detection.
The parameters of the methods were selected in such a way that the number of collision detection queries in each expansion step is the same.
For each conformation and tested scale, the \RA\ method checks $m=100$ candidate configurations for collision, and \RB\ checks $5 \times 20$ configurations for collision.
%collision-detection queries.
Other parameters of the \RB\ planner were the same as for \RA.
The planning was terminated after $\Imax$ iterations or if the tree approached the active site to the distance less than 2~\AA\ (i.e., the 
geometric center of the ligand was closer than 2~\AA\ from the active site).

@@

The traversability rate is shown in Tab.~\ref{tab::main}. 
Only the traversability of the main tunnel is shown, as no trajectory was found in the side tunnel for most of the tested ligands.
The only exception is the ligand m037t, which was the smallest tested one (consisting of 8 atoms), and therefore it could fit even inside the side tunnel.
The traversability rate is higher for the most scaled-down ligands ($\smin=0.5$) and it decreases with the increasing scale.
For the DCP ligand, the trajectories were detected in 18~\% of frames for scale $\smin=0.8$ and the traversability of this ligand
to the active site was approved by a separate MD simulation~\cite{marques2017catalytic}.
%For the scale $\smin=0.8$, the most traversable frames were detected for ligands DCP (18~\%), m037t (45~\%) and m038t (30~\%).
%The traversability of the DCP ligand to active site was approved by a separate MD simulation~\cite{marques2017catalytic}.
Based on the success rate, also other ligands (e.g., m037t with traversability rate 45~\% at $\smin$=0.8) seem to be promising.
However, their binding was not yet approved by dedicated MD simulations.
The planning was realized with 50 conformations of the ligands, but not all of them were used, which is shown in Tab.~\ref{tab::m040c}.
%The selection of 50 conformations was satisfactory.

@@

\begin{table}[t]
\caption{\label{tab::m040c}
\small
Examples of elongated/packed conformation for m040 (1,5-Dichloropentane).
The table shows the average number of different ligand conformations (out of 50) used in trajectories reaching the active site in the 4E46 protein.
}
\centering
{\footnotesize
%\def\arraystretch{0.9}
\begin{tabular}{cc}
\rotatebox{0}{\hskip 5pt Long} \\ \includegraphics[width=0.07\textwidth]{fig/m040-conf1} \\
\rotatebox{0}{Packed} \\ \includegraphics[width=0.0375\textwidth]{fig/m040-conf2}
\end{tabular}
}
{\scriptsize
\def\arraystretch{0.9}
\input usedc.tex
}
\end{table}

@@

%The case was the analysis of Leukotriene A4 hydrolase (PDB ID 4L2L).
The task of the second analysis was to find the trajectory for the ligand 4-(4-benzylphenyl)-1,3-thiazol-2-amine into the active
site of A4 hydrolase (PDB ID 4L2L, 9666 atoms).
%This ligand is a known inhibitor of the protein 4L2L.
%Based on the location of inhibitor inside the protein we assume that the inhibitor is able 
The tCONCOORD tool~\cite{seeliger2007geometry} was used to generate 1000 conformations of the protein 4L2L.
The initial PDB structure was minimized by GROMACS 4.0.7
%\cite{hess2008gromacs} 
 using the input parameters recommended on the example page of tCONCOORD~\cite{tconcoord}.
The protein conformations were generated using the default settings for tCONCOORD in the PERT mode. % which only perturbs the starting configuration of atoms instead of complete randomization.
The tunnels leading to the active site defined by residues 134 (Alanine) and 311 (Phenylalanine) were computed by CAVER 3.0, using the probe 0.9~\AA.
All 1000 conformations contained at least 5 tunnels, more than 900 conformations contained 9 tunnels.
The ordering of the detected tunnels slightly differs in different conformations.
Therefore, the ligand traversability towards the active site was computed for the first three widest tunnels (with average bottlenecks 1.22~\AA, 1.20~\AA\, and 1.19~\AA\ with the average lengths 22.3~\AA, 25.3~\AA\, and 28.4~\AA, respectively).
The conformations of the ligand were prepared in the same way as in the previous section.
From the available $|\L|=736$ low-energy conformations, 100 were selected using the `S' strategy.

@@

As the tested ligand is a known inhibitor of the protein 4L2L~\cite{marques2017enzyme} we assumed that the ligand can reach the active site.
Due to large size of the ligand (33 atoms), low scale-down factors $\smin\in\{0.4,0.5,0.6\}$ were used and
the active site was considered as reached if the distance to the geometric center of the ligand was less than 5~\AA.
All parameters of the \RA\ were set as in the previous experiment, except $\Imax=100\cdot10^3$.
The proposed planner was able to find trajectories even for this large ligand inside the long tunnels, which is
shown by the traversability rates (Tab.~\ref{tab::4l2l}).
Example of the solution is depicted in Fig.~\ref{fig::inhibitor}.
